---
title: "Power Analysis"
output:
  word_document:
    toc: yes
    toc_depth: '3'
  html_document:
    code_folding: hide
    toc: yes
    toc_depth: 3
    number_sections: yes
    toc_float: yes
---

# Notes

* Feb 18, 2021: I have only run models assuming between-cluster variance of about 10%. I am guessing that this is a high estimate. Also, I have assumed that we will have all 300 participants in both arms with the outcome variable not missing. Future analyses should consider other estimates of between-cluster variance and should evaluate the potential impact of attrition. 
* Feb 21, 2021: I added a simulation estimating power for the GEE model in which reSET-O yields a reduction in the OR of 40% with only 5% of the total variance between cluster (as opposed to the 10% I had assumed before). With a 5% between-cluster variance, we could expect to detect a 40% reduction in odds with approximately $1-\beta=.80$, the target power. 

# Prepare workspace
Load the required packages.
```{r warning=F, message=FALSE}
require( clusterPower )
require( powerSurvEpi )
```


# GEE Models

## 10% Between-Cluster Variance

Each of the models in this section assume that the between-cluster variance accounts for 10% of the total variance within both the reSET-O and TAU groups. 

### GEE Model Assuming OR=0.50
Estimate power for rejecting the null hypothesis that difference in the odds of non-abstinence (0=abstinent; 1=used drugs) across the reSET-O and treatment as usual care (TAU) conditions is 0 at weeks 9-12 of the study. The model assumes the following:

* Random allocation of half the total cluster (k=6) to reSET-O and half (k=6) to TAU
* Equal cluster sizes of 50 patients (300 patients in each group)
* Probability of non-abstinence in TAU assumed to be 40% based on [Maricich et al., 2020](https://www.tandfonline.com/doi/full/10.1080/03007995.2020.1846022)
* Probability of non-abstinence in reSET-O assumed to be 25% based on [Maricich et al., 2020](https://www.tandfonline.com/doi/full/10.1080/03007995.2020.1846022)
* This comes to an assumed odds ratio (OR) = 0.5, a 50% reduction in odds.

```{r warning=F, message=F}
# geesim1 <- cps.binary( nsim = 1000,
#                        nsubjects = 50, # 50 participants per cluster assuming equal sizes
#                        nclusters = 6, # 6 clusters per treatment arm
#                        p1=.40, # 40% non-abstinence rate in TAU in Maricich et al. (2020)
#                        p2=.25, # 25% non-abstinence rate in reSET-0 group in Maricich et al. (2020)
#                        sigma_b_sq=0.024, # variance = p(1-p)=0.24; assume 10% variance is between)
#                        sigma_b_sq2=0.019, # variance = p(1-p)=0.1875; assume 10% variance is between)
#                        alpha=.05,
#                        method='gee',
#                        quiet=F,
#                        seed=0218211,
#                        lowPowerOverride = TRUE)
load( 'geesim1.RData' )
geesim1["power"]                     
```

This simulation suggests that our power would be between 0.94-0.97 -- i.e., well powered.  

### GEE Model Assuming OR=0.55

Next, we will calculate power using a simulation assuming an in-between effect where reSET-O results in a ~45% reduction in the odds.

* Assume a 40% non-abstinence rate in TAU
* Assume a 27% non-abstinence rate in reSET-O
* Keep all of the rest of the assumptions the same as in the previous simulation

```{r warning=F, message=F}
# geesim5 <- cps.binary( nsim = 1000,
#                        nsubjects = 50, # 50 participants per cluster assuming equal sizes
#                        nclusters = 6, # 6 clusters per treatment arm
#                        p1=.40, # 40% non-abstinence rate in TAU in Maricich et al. (2020)
#                        p2=.27, # 27% non-abstinence rate in reSET-0 group 
#                        sigma_b_sq=0.024, # variance = p(1-p)=0.24; assume 10% variance is between)
#                        sigma_b_sq2=0.022, # variance = p(1-p)=0.21; assume 10% variance is between)
#                        alpha=.05,
#                        method='gee',
#                        quiet=F,
#                        seed=0218211,
#                        lowPowerOverride = TRUE )
load( 'geesim5.RData' )
geesim5["power"]
```

This simulation suggests that our power would be well-powered 0.85-0.89 to detect an OR=0.55. 

### GEE Model Assuming OR=0.60

Next, we will calculate power using a simulation assuming an in-between effect where reSET-O results in a ~40% reduction in the odds.

* Assume a 40% non-abstinence rate in TAU
* Assume a 29% non-abstinence rate in reSET-O
* Keep all of the rest of the assumptions the same as in the previous simulation

```{r warning=F, message=F}
# geesim4 <- cps.binary( nsim = 1000,
#                        nsubjects = 50, # 50 participants per cluster assuming equal sizes
#                        nclusters = 6, # 6 clusters per treatment arm
#                        p1=.40, # 40% non-abstinence rate in TAU in Maricich et al. (2020)
#                        p2=.29, # 29% non-abstinence rate 
#                        sigma_b_sq=0.024, # variance = p(1-p)=0.24; assume 10% variance is between)
#                        sigma_b_sq2=0.021, # variance = p(1-p)=0.21; assume 10% variance is between)
#                        alpha=.05,
#                        method='gee',
#                        quiet=F,
#                        seed=0218211,
#                        lowPowerOverride = TRUE )
load( 'geesim4.RData' )
geesim4["power"]
```

This simulation suggests that our power would be between 0.74-0.79 -- i.e., approaching the goal of 0.80. 


### GEE Model Assuming OR=0.65
Now we will calculate power using a simulation assuming an in-between effect where reSET-O results in a ~35% reduction in the odds.

* Assume a 40% non-abstinence rate in TAU
* Assume a 30% non-abstinence rate in reSET-O
* Keep all of the rest of the assumptions the same as in the previous simulation

```{r warning=F, message=F}
# geesim3 <- cps.binary( nsim = 1000,
#                        nsubjects = 50, # 50 participants per cluster assuming equal sizes
#                        nclusters = 6, # 6 clusters per treatment arm
#                        p1=.40, # 40% non-abstinence rate in TAU in Maricich et al. (2020)
#                        p2=.30, # 30% non-abstinence rate
#                        sigma_b_sq=0.024, # variance = p(1-p)=0.24; assume 10% variance is between)
#                        sigma_b_sq2=0.021, # variance = p(1-p)=0.21; assume 10% variance is between)
#                        alpha=.05,
#                        method='gee',
#                        quiet=F,
#                        seed=0218211,
#                        lowPowerOverride = TRUE )
load( 'geesim3.RData' )
geesim3["power"]         
```

This simulation suggests that our power would be between 0.66-0.71 -- i.e., moderately powered.

### GEE Model Assuming OR=0.80
Now we will calculate power using a simulation assuming a more modest effect where reSET-O results in a 20% reduction in the odds.

* Assume a 40% non-abstinence rate in TAU
* Assume a 35% non-abstinence rate in reSET-O
* Keep all of the rest of the assumptions the same as in the previous simulation

```{r warning=F, message=F}
# geesim2 <- cps.binary( nsim = 1000,
#                        nsubjects = 50, # 50 participants per cluster assuming equal sizes
#                        nclusters = 6, # 6 clusters per treatment arm
#                        p1=.40, # 40% non-abstinence rate in TAU in Maricich et al. (2020)
#                        p2=.35, # 35% non-abstinence rate 
#                        sigma_b_sq=0.024, # variance = p(1-p)=0.24; assume 10% variance is between)
#                        sigma_b_sq2=0.023, # variance = p(1-p)=0.2275; assume 10% variance is between)
#                        alpha=.05,
#                        method='gee',
#                        quiet=F,
#                        seed=0218211,
#                        lowPowerOverride = TRUE )
load( 'geesim2.RData' )
geesim2["power"]                    
```

This simulation suggests that our power would be between 0.23-0.29 -- i.e., poorly powered.  

## 5% Between-Cluster Variance
In this section, we rerun the simulation models that fell short of the desired power ($\beta-1=.80$) with 10% between-cluster variability assuming just 5% between-cluster variability, which seems more probably given the design. 


### 
Rerun geesim4, which assumes a ~40% reduction in the odds but with only 5% of the total variance between clusters. The geesim4 model assuming 10% between-cluter variance yielded a power estimate (0.74-0.79) that fell just short of the goal.

* Assume a 40% non-abstinence rate in TAU
* Assume a 29% non-abstinence rate in reSET-O


```{r warning=F, message=F}
# geesim6 <- cps.binary( nsim = 1000,
#                        nsubjects = 50, # 50 participants per cluster assuming equal sizes
#                        nclusters = 6, # 6 clusters per treatment arm
#                        p1=.40, # 40% non-abstinence rate in TAU in Maricich et al. (2020)
#                        p2=.29, # 29% non-abstinence rate in reSET-0
#                        sigma_b_sq=0.012, # variance = p(1-p)=0.24; assume 5% variance is between)
#                        sigma_b_sq2=0.010, # variance = p(1-p)=0.21; assume 5% variance is between)
#                        alpha=.05,
#                        method='gee',
#                        quiet=F,
#                        seed=0221211,
#                        lowPowerOverride = TRUE )
load( 'geesim6.RData' )
geesim6["power"]
```

This simulation suggests that our power would be between 0.74-0.79 -- i.e., approaching the goal of 0.80. 



```{r}

```



# Cox Propotional Hazards Models

## Non-Abstinence Time-To-Onset
These models should be taken with a **big** grain of salt. Power analysis for a proportional hazards model with individuals clustered within groups is a bit beyond my capabilities. I believe this would require simulations that are beyond my expertise. I ran a number of quick-and-dirty models varying the probability of an event (drug use) over the full follow-up period in the reSET-O group and the assumed hazard ratio. For all models, I assumed that 40% of TAU participants would have an event over the full follow-up. As with the previous examples, I am assuming 300 participants per treatment arm ($N=600$).

The consequence of failing to account for clustering is having too small standard error estimates. I used a conservative $\alpha = .025$ to try to counteract this -- a terribly unsophisticated approach. The [powerSurvEpi package](https://cran.r-project.org/web/packages/powerSurvEpi/powerSurvEpi.pdf) was used to obtain these estimates.

```{r warning=F, message=F}
surv.80 <- powerCT.default(nE = 300,
                           nC = 300,
                           pE = 0.35,
                           pC = 0.40,
                           RR = 0.8,
                           alpha = 0.025 )
surv.75 <- powerCT.default(nE = 300,
                           nC = 300,
                           pE = 0.33,
                           pC = 0.40,
                           RR = 0.75,
                           alpha = 0.025 )
surv.70 <- powerCT.default(nE = 300,
                           nC = 300,
                           pE = 0.32,
                           pC = 0.40,
                           RR = 0.7,
                           alpha = 0.025 )
surv.65 <- powerCT.default(nE = 300,
                           nC = 300,
                           pE = 0.30,
                           pC = 0.40,
                           RR = 0.65,
                           alpha = 0.025 )
surv.60 <- powerCT.default(nE = 300,
                           nC = 300,
                           pE = 0.29,
                           pC = 0.40,
                           RR = 0.60,
                           alpha = 0.025 )

plot( sort( seq(.80,.60,-.05), decreasing = T ),
      c(surv.80[1], surv.75[1], surv.70[1], surv.65[1], surv.60[1]),
      type='b',
      ylim = c(0,1),
      lwd=2,
      col=2,
      ylab='Estimated Power',
      xlab='Hazard Ratio',
      main='Proportional Hazards Model',
      sub='40% Probability of Event in Control over Full Time Period (2-tailed alpha=.025)',
      font.lab=2)
abline( h = .8, col = 'darkblue', lty = 3 )
abline( v= .65, col = 'darkblue', lty = 3 )
```

As can be seen in the plot above, these very unsophisticated power estimates suggest that we could detect a hazard ratio of approximately $.65-.60$ to achieve $1-\beta=.80$. 

## Time to Dropout

```{r warning=F, message=F}
drop.50 <- powerCT.default(nE = 300,
                           nC = 300,
                           pE = 0.18,
                           pC = 0.32,
                           RR = 0.5,
                           alpha = 0.025 )
drop.55 <- powerCT.default(nE = 300,
                           nC = 300,
                           pE = 0.18,
                           pC = 0.32,
                           RR = 0.55,
                           alpha = 0.025 )
drop.60 <- powerCT.default(nE = 300,
                           nC = 300,
                           pE = 0.19,
                           pC = 0.32,
                           RR = 0.6,
                           alpha = 0.025 )
drop.65 <- powerCT.default(nE = 300,
                           nC = 300,
                           pE = 0.21,
                           pC = 0.32,
                           RR = 0.65,
                           alpha = 0.025 )
drop.70 <- powerCT.default(nE = 300,
                           nC = 300,
                           pE = 0.225,
                           pC = 0.32,
                           RR = 0.70,
                           alpha = 0.025 )

Power <- c(surv.60,surv.65,surv.70,surv.75,surv.80,drop.50,drop.55,drop.60,drop.65,drop.70)
HazardRatio <- c( seq(.6, .8, .05), seq(.5, .7, .05) )
Model <- rep( c('Time To Non-Abstinence','Time To Dropout'), each = 5 )
tto <- data.frame( Model, HazardRatio, Power ) 

require(ggplot2)

ttoplot <- ggplot( tto, aes( HazardRatio, Power ) ) +
  geom_point(size=3) +
  facet_wrap( ~Model ) +
  geom_hline( yintercept = .80, color='red', linetype = 'dashed' ) +
  ggtitle( label='Power: Proportional Hazards Models', subtitle = '2-tailed alpha = .025' ) +
  theme( axis.title = element_text( face='bold', size = 14 ),
         plot.title = element_text( face='bold', size = 16, color = 'darkblue' ),
         strip.text = element_text( size = 14, face = 'bold'),
         axis.text = element_text( size=12 ),
         plot.subtitle = element_text(size=12, face='bold'))
ttoplot

```